KIN: Method to infer relatedness using low coverage ancient DNA

### 2022-03-21 paper_overleaf/main.tex

\title{Method to infer relatedness using low coverage ancient DNA}
\author{Divya Ratan Popli, Benjamin M. Peter}
\date{5 August 2021}
\linenumbers

@@

\noindent Knowledge of familial relationships is an important part of several research fields. However, estimating relatedness from ancient DNA can be be difficult since DNA sequences extracted from ancient bones may have low coverage, ascertainment bias (if the sequences have been captured with DNA probes), and contamination from modern humans. In addition, the population of interest may have long runs of homozygosity (ROH) due to recent inbreeding or/and small population size. This can affect the relatedness estimates. Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and degree of relatedness up to 3rd degree, while additionally differentiating between siblings and parent-child. The output can be further refined with another HMM that we developed to estimate ROH proportions.
\end{abstract}

@@

In ancient DNA, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an  understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_KINship-based_2019,sikora_ancient_2017}. For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals in the mating animals is important to ensure diversity. ~\cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}

@@

There are numerous methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to estimate the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}. PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from observed IBS, using the allele frequencies at each SNP in samples \cite{purcell_plink_2007}. KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}. .

@@

\subsection{Problems with ancient data, and methods that address these issues}
One major issue with applying the above mentioned methods to ancient DNA is that, frequently, the amount of endogenous DNA is very low, making it very difficult to get genotype calls.
Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}.In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods. However, these approaches typically still require at least 2x coverage, since genotype likelihoods may not be very informative at lower coverages. Apart from the low amount of endogenous DNA sequences present, ancient DNA analyses often have additional challenges like contamination from modern populations \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in case of captured sequences. In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations. However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}. Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4x$ coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}. In this study, we show that contamination from modern humans can cause some of the individuals in ancient populations to look more unrelated than they are. This happens because of comparison of endogenous reads to contaminating reads which would give higher differences than comparison of endogenous reads from individuals from ancient population. It is worth mentioning that different DNA libraries may have different proportion of contamination levels, even when they have been sampled from the same specimen. Hence, contamination can reduce the power to detect related pairs, especially when the target population is quite diverged from contaminating population(s). In addition, the target populations may have long runs of homozygosity (ROH) due to a small population size, or recent inbreeding. Long ROH causes related individuals to seem genetically closer to each other, while does not affect the genetic distance between unrelated individuals. This effect may cause an increase in false positives. Moreover, ancient DNA is commonly captured with a SNP array to enrich for endogenous sequences. The way that these SNPs are defined (e.g. selecting SNPs variable in certain individuals) becomes relevant, because relatedness methods based on the fraction of sites in different IBS states are highly sensitive to ascertainment bias. KING robust, for example, differentiates between different relatedness by plotting KING‐robust KINship against fraction of sites in IBS0 \cite{manichaikul_robust_2010}. Another method from Rosenberg,2006 \cite{rosenberg_standardized_2006} uses the fraction of sites in all three IBS states to classify related individuals in the HGDP-CEPH Human Genome Diversity Cell Line Pane. These methods may give different results when applied to data with different ascertainment schemes \cite{waples_allele_2019}.

@@

\subsubsection{Relatedness cases}
We run our model for the following relatedness cases: Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, Identical.
We estimated the transition matrix corresponding to each relatedness case using simulations (see section XX). While testing KIN on simulations, we found that classification performance for $4^{th}$ and $5^{th}$ degree relatives is poor, and we have low power to distinguish them from unrelated pairs of individuals. However, we would like to classify 4-th and 5-th degree relatives as unrelated, and so we include these cases in the algorithm. Similarly, we observe that we have low power to differentiate between half-siblings, avuncular and grandparent-grandchild, and we merge these classifications as $2^{nd}$ degree in the final results. We report the final pairwise classification in the following categories: Unrelated, Third Degree, Second Degree, Parent-Child, Siblings, Identical individuals. We output the relatedness corresponding to the maximum likelihood model, and the confidence as given by the log likelihood ratio between the two highest likelihood models.

@@

\subsubsection{Contamination}
Contamination by present-day people is a common feature in ancient DNA \cite{peyregne_present-day_2020}. Even low levels of contamination may make samples look less similar, and thus distort relatedness analyses. This issue is particularly pronounced  when analyzing populations with low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations (Skov et al.,2022 (in press)). To address this issue, we develop a heuristic that adjusts both $D_w$ and $N_w$ to remove contamination from the data.

We conceptualize contamination in the following way. If data from individual $i$ has a known contamination rate of $C_i$, this means that any particular read drawn from this individual will actually come from that individual with probability $1-C_i$. With probability $C_i$ the read comes from a contaminant. Thus, when comparing two individuals $i$ and $j$, there are three possible cases: First, with probability $(1-C_i)(1-C_j)$, we compare endogenous reads, which we can use to infer relatedness. With probability $C_i(1-C_j) + C_j(1-C_i)$ we compare an endogenous read from an individual with a contaminant read from the other individual. Finally, with probability $C_iC_j$ we compare contaminant reads from both the samples. If we make the simplifying assumption that both $C_i$ and $C_j$ are small, the terms of the order of $C^2$ can be ignored. Then, we compare contaminant read to endogenous read with the probability $C_{ij} = C_i + C_j$.

Our goal then, is to estimate an expectation of differences from comparison of endogenous reads $\mathbb{E}[D_w^{'}]$, and the total number of sites with overlapping endogenous reads $\mathbb{E}[N_w^{'}]$. To estimate $\mathbb{E}[D_w^{'}]$, we first need to calculate $P(c=0|d=1)$. Here, $c$ is a binary variable with value $0$ or $1$ denoting that the comparison at a site is between endogenous reads, or between endogenous and contaminant reads respectively. Similarly $d$ is a binary variable taking values 0 and 1 for seeing no difference, or a difference, when comparing the reads at a given site. We can further expand this probability:

@@

Here, $P(d=1|c=0)$, referred later to as $\rho_{ij}$ for simplicity, is equivalent to the proportion of differences when we compare endogenous reads from individuals $i$ and $j$. This can be estimated by considering that proportion of pairwise differences ($\frac{\sum_w D_w}{\sum_w N_w}$), referred to as $o_{ij}$ later, can be expressed as a sum of proportion of differences between endogenous reads and that between endogenous and contaminant reads \cite{peyregne_present-day_2020}. 
\begin{align}\label{eq:13}
    o_{i,j} = (1-C_{ij}) \rho_{ij} + C_{ij} \phi \nonumber\\
    or, \rho_{ij} = \frac{o_{ij} - C_{ij} \phi} {1 - C_{ij}}
\end{align}

Here, $\phi$ is the probability of seeing a difference when endogenous read is compared to a contaminant read, and is equivalent to the average divergence between target and contaminating population.
Given this information, we can rewrite eq. \ref{eq:12a}:
\begin{align}\label{eq:12}
    P(c=0|d=1) &= \frac{(1 - C_{ij}) \rho_{ij}}{(1 - C_{ij}) \rho_{ij} + C_{ij} \phi} \nonumber\\
\end{align}
We can further write $\mathbb{E}[D_w^{'}]$ as $D_w P(c=0|d=1)$:
\begin{align}
    \mathbb{E}[D_w^{'}] &= D_w\frac{(1-C_{ij})\rho_{ij}}{(1-C_{ij}) \rho_{ij} + C_{ij} \phi}\\
\end{align}

Similarly, we derive $\mathbb{E}[S_w^{'}]$ as the estimate for number of sites that do not show pairwise difference.
\begin{align}
    \mathbb{E}[S_w^{'}] &= (N_w-D_w) P(e=1|d=0) = (N_w - D_w)\frac{(1-\rho_{ij})(1-C_{ij})}{(1-\rho_{ij})(1-C_{ij}) + C_{ij}(1-\phi)}\\
\end{align}

@@

\paragraph{Estimating transition matrices}
KIn requires a transition matrix for each relatedness case. For the simplest cases like grandparent-grandchild and siblings, it is possible to calculate the transition matrix from first principles, but for most cases this is difficult  because the relatedness cases include multiple cases, or they are non-Markovian, and so  and so for these cases we obtain transition matrices from our simulations.

@@

Apart from the related and unrelated individuals in Pop1, we simulated more haploids in three other populations to create scenarios with ascertainment and contamination (fig. \ref{figS10:pedigree}). We simulated two haploids to form an individual each from two other populations (Pop2, Pop3) with split time of 3500 and 4500 generations with Pop1, and sampling time of 4000 generations and 2000 generations ago respectively. We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from Pop1. We test out the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity with a Markov chain using transition matrix shown below:

@@

From the steps described above, we got genotypes of 17 individuals in Pop1 in presence/absence of ROH and ascertainment. We further simulated 10 haploids (5 diploid individuals) from Pop4 with split time of 20,000 generations with pop1, sampled from the present time. We calculated the allele frequency for Pop4 at each site, and used it to introduce contamination (with fixed contamination percentage for each individual ($<5\%$) in samples from pop1. We generated reads (derived/ancestral) for different genomic coverages ranging from 4x to 0.03x at each position for each individual. We specifically generated reads for four different cases: Ancestral-Endogenous, Ancestral-Contaminant, Derived-Endogenous, Derived-Contaminant. The number of reads for each case was sampled from a Poisson distribution with $\lambda$ calculated from the coverage $\zeta$, allele frequency or genotype (G $\in$ 0,0.5,1), and contamination estimate:

@@

@@

\paragraph{Motivating Example}
We predict different categories of relatedness with KIN, and test it's performance on simulated pedigrees. We use these simulations to find the limits of our method, and compare it to READ, a widely used method. We introduce cases of contamination, ascertainment, low coverages, and ROH to compare the two methods. Finally we apply KIN to published data and show that it can give reliable information about genetic kinship, that may not be available from other available methods. We show the output of different Relatedness HMMs in Fig.\ref{fig1:ibd}, when they are applied to a pair of siblings.

@@

\paragraph{Evaluating the ROH detection}
We applied the ROH-HMM and relatedness HMM on 60 runs of simulations, to get relatedness estimates, and compared the final relatedness estimates to the true relatedness for different pairs at coverages ranging from 4x to 0.05x. 
Performance of ROH-HMM under scenarios of ascertainment bias, contamination and long ROH (described in section Simulations), and for different coverages is shown in Fig.\ref{fig2:ROH}. We observe that as coverage decreases, the raw heterozygosity estimates become more noisy. In the case of no ROH in the data (Fig \ref{fig2:ROH}A, we correctly infer that there is no ROH for coverage of 0.2x and 4x, but at the lowest coverage of 0.1x there are few regions where we erroneously call some ROH. In all three scenarios with ROH ((Fig \ref{fig2:ROH}), the model prediction deviates from the true ROH probability by 8-12 \%) for the three coverages. Hence, we do not see any large adverse effects of ascertainment bias and contamination on the performance of ROH HMM.

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot.png}
    \centering
    \caption{Estimation of ROH probabilities along the genome in simulations with different cases of ROH, contamination, and ascertainment. Top panel shows proportion of differences in a simulated individual along the genome, and the bottom panel shows probability of seeing no ROH. (A) Simulation with no ascertainment, contamination, or ROH. (B) Simulation with ROH. (C) Simulation with ROH and ascertainment. (D) Simulation with ROH and contamination.}
    \label{fig2:ROH}
\end{figure}

\paragraph{Critical Values}
The log likelihood-ratio ($\Lambda$) is commonly used in likelihood-ratio tests to compare nested models. However, in our cases, models are not nested, and so standard likelihood-ratio theory does not apply. For this reason, we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values. We plotted the true positive and false positive rates with a cutoff on the log likelihood ratio in Fig.\ref{fig3:cutoff}. Here, and in Fig.\ref{fig4:Comparison_READ_KIN}, we have two relatedness labels: 'Unrelated' and 'Unrelated w/o 3rd Degree'. Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated. This is useful for comparison to READ, since READ classifies relatedness up to Second Degree. 'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is considered as separate from the unrelated individuals. Fig.\ref{fig3:cutoff} shows that for all relatedness cases except for $3^{rd}$ Degree, the false positive rate is below $5\%$ even when no cut off is used. We observe that for $3^{rd}$ Degree relation, using a cutoff of 1 brings down the false positive rate close to $5\%$ for all coverages except 0.05x. In conclusion, it is recommended to use a cutoff of 1 for most cases.

@@

We use a cutoff of 1 to compare our method performance to READ, while using a cutoff of 1 standard deviation for READ. Fig.\ref{fig4:Comparison_READ_KIN} shows this comparison using false positive and true positive rates for both the methods at different coverages under different simulations. We show that both methods have similar performance in the control case, and cases with ROH and ascertainment, except for $2^{nd}$ Degree relation, where KIN has higher power, especially at low coverages. We show that adding contamination adversely affects both the False Positve and True Positive rates corresponding to 1st and 2nd Degree relations for READ, while this has no effect on KIN performance.

@@

\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/plotimg/fil0_relatable_plot.png}
    \centering
    \caption{Application of KIN on Neandertal specimens from Chagyrskaya. The color of a square represents the relatedness, while the number within denotes log likelihood ratio between the two maximum likelihood models.}
    \label{fig5:Chagyrskaya_KIN}
\end{figure}

We further applied KIN to a genome-wide dataset of 118 ancient individuals sequenced and analyzed by Mittnik et al.,2019. We compared our relatedness estimates to those obtained from the authors using READ and lcMLKIN, and found that in most cases the three methods agree. Out of 6903 pairs, 346 pairs show different relatedness for KIN and READ (table S2). However, there are only 31 pairs where KIN has $\Lambda >1$. We looked in detail at some of these pairs, and we found that it is possible to deduce the correct relatedness for several of these pairs by plotting the proportion of differences along the genome. In all the pairs that we analyzed in this way, KIN seems to predict the true relatedness. We show few examples of such cases in fig. S\ref{figS8:eg1}. We found 20 pairs in this subset of 346 pairs where lcMLKIN has sufficient data (Overlapping SNPs $>$ 10000). Among these 20 pairs, we found that lcMLKIN matches KIN estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLKIN predicts 3rd-5th degree, KIN predicts second degree) and pair AITI72-AITI77A (READ and lcMLKIN predict unrelated, while KIN estimates 3rd degree with log likelihood=1.7).  
We found 43 pairs for which READ estimates first degree, and compared the estimates of lcMLKIN and KIN for these cases (suppl table3). There were a total of 20 pairs where lcMLKIN had sufficient data, and 26 pairs where $loglikelihood ratio>2$ for KIN. Among the 19 overlapping pairs, we found four pairs where KIN and lcMLKIN differ, and a visual inspection shows that in these cases, KIN seems to have the correct estimate (fig. S\ref{figS9:eg2}).

@@

>>>>

\title{KIN: Method to infer relatedness using low coverage ancient DNA}
\author{Divya Ratan Popli, Benjamin M. Peter}
\date{5 August 2021}
\linenumbers

@@

\noindent Knowledge of familial relationships is an important part of several research fields. However, estimating relatedness from ancient DNA can be difficult since DNA sequences extracted from ancient remains may have low coverage, ascertainment bias (if the sequences have been captured with DNA probes), and contamination from modern human DNA. In addition, the population of interest may have long runs of homozygosity (ROH) due to recent inbreeding or/and small population size. This can affect the relatedness estimates. Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and degree of relatedness up to 3rd degree, while additionally differentiating between siblings and parent-child. The output can be further refined with another HMM that we developed to estimate ROH proportions.
\end{abstract}

@@

In ancient DNA studies, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_KINship-based_2019,sikora_ancient_2017}. For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to ensure diversity. ~\cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}

@@

There are numerous methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to estimate the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}. PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from observed IBS, using the allele frequencies at each Single Nucleotide Polymorphism (SNP) in the analyzed samples \cite{purcell_plink_2007}. KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies in the samples (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}. .

@@

\subsection{Problems with ancient data, and methods that address these issues}
One major issue with applying the above mentioned methods to ancient DNA is that, frequently, the amount of endogenous DNA is very low, making it very difficult to get genotype calls.
Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}.In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods. However, these approaches typically still require at least 2x coverage, since genotype likelihoods may not be very informative at lower coverages. Apart from the low amount of endogenous DNA sequences present, ancient DNA analyses often have additional challenges like contamination from present-day DNA \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in case of captured sequences. In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations. However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}. Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4x$ coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}. In this study, we show that contamination can cause some of the individuals in ancient populations to look more unrelated than they are. This happens because of comparison of endogenous reads to contaminating reads which yields more differences than comparison of endogenous reads from the ancient individuals. It is worth mentioning that different DNA libraries may have different levels of contamination, even when they have been sampled from the same specimen. Hence, contamination can reduce the power to detect related pairs, especially when the target population is quite diverged from contaminating population(s). In addition, the genomes of target individuals may have long runs of homozygosity (ROH) due to a small population size, or recent inbreeding. Long ROH causes related individuals to seem genetically closer to each other, while does not affect the genetic distance between unrelated individuals. This effect may cause an increase in false positives. Moreover, ancient DNA is commonly captured with a SNP array to enrich for endogenous sequences. The way that these SNPs are defined (e.g. selecting SNPs variable in certain individuals) becomes relevant, because relatedness methods based on the fraction of sites in different IBS states are highly sensitive to ascertainment bias. KING robust, for example, differentiates between different relatedness by plotting KING‐robust KINship against fraction of sites in IBS0 \cite{manichaikul_robust_2010}. Another method from Rosenberg,2006 \cite{rosenberg_standardized_2006} uses the fraction of sites in all three IBS states to classify related individuals in the HGDP-CEPH Human Genome Diversity Cell Line Panel. These methods may give different results when applied to data with different ascertainment schemes \cite{waples_allele_2019}.

@@

\subsubsection{Relatedness cases}
We run our model for the following relatedness cases: Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, Identical.
We estimated the transition matrix corresponding to each relatedness case using simulations (see section XX). While testing KIN on simulations, we found that classification performance for $4^{th}$ and $5^{th}$ degree relatives is poor, and we have low power to distinguish them from unrelated pairs of individuals. However, we would like to classify 4-th and 5-th degree relatives as unrelated, and so we include these cases in the algorithm. Similarly, we observe that we have low power to differentiate between half-siblings, avuncular and grandparent-grandchild, and we merge these classifications as $2^{nd}$ degree in the final results. We report the final pairwise classification in the following categories: Unrelated, Third Degree, Second Degree, Parent-Child, Siblings, Identical individuals. We output the relatedness corresponding to the maximum likelihood model, and the confidence as given by the log likelihood ratio between the two highest likelihood models. The log likelihood-ratio is commonly used in likelihood-ratio tests to compare nested models. However, in our cases, models are not nested, and so standard likelihood-ratio theory does not apply. For this reason, we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig \ref{fig3:cutoff}).

@@

\subsubsection{Contamination}
Contamination by the DNA from present-day people is a common feature in ancient DNA \cite{peyregne_present-day_2020}. Even low levels of contamination may make samples look less similar, and thus distort relatedness analyses. This issue is particularly pronounced  when analyzing populations with low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations (Skov et al.,2022 (in press)). To address this issue, we develop a heuristic that adjusts both $D_w$ and $N_w$ to remove contamination from the data.

We conceptualize contamination in the following way. If data from individual $i$ has a known contamination rate of $C_i$, this means that any particular read drawn from this individual will actually come from that individual with probability $1-C_i$. With probability $C_i$ the read comes from a contaminant. Thus, when comparing two reads, one drawn from individual $i$ and the other drawn from individual $j$, there are three possible cases: First, with probability $(1-C_i)(1-C_j)$, both reads are endogenous and we can use them to infer relatedness. With probability $C_i(1-C_j) + C_j(1-C_i)$, one read is endogenous and the other is contaminant. Finally, with probability $C_iC_j$, both reads are contaminant. If we make the simplifying assumption that both $C_i$ and $C_j$ are small, the terms of the order of $C^2$ can be ignored. Then, we compare contaminant read to endogenous read with the probability $C_{ij} = C_i + C_j$.

Given that we know the level of contamination (estimated from other methods),our goal then, is to estimate the expected number of differences from comparison of endogenous reads $\mathbb{E}[D_w^{'}]$, and the total number of sites with overlapping endogenous reads $\mathbb{E}[N_w^{'}]$. To estimate $\mathbb{E}[D_w^{'}]$, we first need to calculate $P(c=0|d=1)$. Here, $c$ is a binary variable with value $0$ or $1$ denoting that the comparison at a site is between endogenous reads, or between endogenous and contaminant reads respectively. Similarly $d$ is a binary variable taking values 0 and 1 for seeing no difference, or a difference, when comparing the reads at a given site. We can further expand this probability:

@@

Here, $P(d=1|c=0)$, referred later to as $\rho_{ij}$ for simplicity, is equivalent to the proportion of differences when we compare endogenous reads from individuals $i$ and $j$. This can be estimated by considering that the proportion of pairwise differences $P(d=1) =\frac{\sum_w D_w}{\sum_w N_w}$, referred to as $o_{ij}$ later, can be expressed as a linear combination of the proportion of differences between endogenous reads and that between endogenous and contaminant reads \cite{peyregne_present-day_2020}. 
\begin{align}\label{eq:13}
    o_{i,j} = (1-C_{ij}) \rho_{ij} + C_{ij} \phi \nonumber\\
    or, \rho_{ij} = \frac{o_{ij} - C_{ij} \phi} {1 - C_{ij}}
\end{align}

Here, $\phi$ is the probability of seeing a difference when an endogenous read is compared to a contaminant read, and is equivalent to the average sequence divergence between individuals of the target and contaminating populations.
Given this information, we can rewrite eq. \ref{eq:12a}:
\begin{align}\label{eq:12}
    P(c=0|d=1) &= \frac{(1 - C_{ij}) \rho_{ij}}{(1 - C_{ij}) \rho_{ij} + C_{ij} \phi} \nonumber\\
\end{align}
We can further write $\mathbb{E}[D_w^{'}]$ as $D_w P(c=0|d=1)$:
\begin{align}
    \mathbb{E}[D_w^{'}] &= D_w\frac{(1-C_{ij})\rho_{ij}}{(1-C_{ij}) \rho_{ij} + C_{ij} \phi}\\
\end{align}

Similarly, we derive $\mathbb{E}[S_w^{'}]$ as the estimate for the number of sites that do not show pairwise differences.
\begin{align}
    \mathbb{E}[S_w^{'}] &= (N_w-D_w) P(e=1|d=0) = (N_w - D_w)\frac{(1-\rho_{ij})(1-C_{ij})}{(1-\rho_{ij})(1-C_{ij}) + C_{ij}(1-\phi)}\\
\end{align}

@@

\paragraph{Estimating transition matrices}
KIN requires a transition matrix for each relatedness case. For the simplest cases like grandparent-grandchild and siblings, it is possible to calculate the transition matrix from first principles, but for most cases this is difficult  because the relatedness cases include multiple cases, or they are non-Markovian, and so for these cases we obtain transition matrices from our simulations.

@@

Apart from the related and unrelated individuals in Pop1, we simulated more haploids in three other populations to create scenarios with ascertainment bias and contamination (fig. \ref{figS10:pedigree}). We simulated two haploids to form an individual each from two other populations (Pop2, Pop3) with split time of 3500 and 4500 generations with Pop1, and sampling time of 4000 generations and 2000 generations ago respectively. We identified the sites that were polymorphic among these two individuals, and used these sites to ascertain the genomes of individuals from Pop1. We test out the performance of our method in presence of long ROH ($\sim17\%$), by simulating regions of homozygosity with a Markov chain using transition matrix shown below:

@@

From the steps described above, we got genotypes of 17 individuals in Pop1 in presence/absence of ROH and ascertainment. We further simulated 10 haploids (5 diploid individuals) from Pop4 with split time of 20,000 generations with pop1, sampled from the present time. We calculated the allele frequency for Pop4 at each site, and used it to introduce contamination (with fixed contamination percentage for each individual ($<5\%$) in samples from Pop1. We generated reads (derived/ancestral) for different genomic coverages ranging from 4x to 0.03x at each position for each individual. We specifically generated reads for four different cases: Ancestral-Endogenous, Ancestral-Contaminant, Derived-Endogenous, Derived-Contaminant. The number of reads for each case was sampled from a Poisson distribution with $\lambda$ calculated from the coverage $\zeta$, allele frequency or genotype (G $\in$ 0,0.5,1), and contamination estimate:

@@

We run simulations with different cases referred to as control (with no ROH, contamination or ascertainment bias), R (with ROH), A (with ascertainment bias), and C (with fixed levels of contamination for each individual).

@@

\paragraph{Motivating Example}
We predict different categories of relatedness with KIN, and test it's performance on simulated pedigrees. We use these simulations to find the limits of KIN as well as ROH HMM. We introduce cases of contamination, ascertainment, ROH, and low coverages to compare our method to READ. Finally we apply KIN to published data and show that it can give reliable information about genetic kinship, that may not be available from other available methods. Here, in Fig.\ref{fig1:ibd} we show the output of different Relatedness HMMs when they are applied to a pair of siblings.

@@

\paragraph{Critical Values}
 We plotted the true positive and false positive rates with a cutoff on the log likelihood ratio in Fig.\ref{fig3:cutoff}. Here, and in Fig.\ref{fig4:Comparison_READ_KIN}, we have two relatedness labels: 'Unrelated' and 'Unrelated w/o 3rd Degree'. Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated. This is useful for comparison to READ, since READ classifies relatedness up to Second Degree. 'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals. Fig.\ref{fig3:cutoff} shows that for all relatedness cases except for $3^{rd}$ Degree, the false positive rate is below $5\%$ even when no cut off is used. We observe that for $3^{rd}$ Degree relation, using a cutoff of 1 brings down the false positive rate close to $5\%$ for all coverages except 0.05x. In conclusion, it is recommended to use a cutoff of 1 for most cases.

@@

In our assessment of critical value for log likelihood ratio, we used the control simulations. However, we test performance of KIN on different simulation cases including case R, which is also based on performance of ROH HMM.

\paragraph{Evaluating the ROH detection}
 Fig.\ref{fig2:ROH} shows the performance of ROH-HMM under cases of control, R, A, C and simulations (described in section Simulations), and for different coverages. We observe that as coverage decreases, the raw heterozygosity estimates become more noisy. In the control case (Fig \ref{fig2:ROH}, we correctly infer that there is no ROH for most of the time, but occasionally we erroneously call some ROH as shown in the figure. We observe that the model predicts ROH probabilities with biases of upto -0.007, -0.081, -0.083, and -0.091 and variances upto 0.006, 0.036, 0.036, and 0.037 in control, R, C, and A cases respectively as shown in supplementary table \ref{tab:Table 2}. The negative bias in all cases shows that on an average, the model over-estimates probability of ROH, and this effect is seen more often at higer coverages. We do not see any large adverse effects of ascertainment bias and contamination on the performance of ROH HMM.

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot_final.png}
    \centering
    \caption{Estimation of ROH probabilities along the genome in simulations with different cases of ROH, contamination, and ascertainment. Top panel shows proportion of differences in a simulated individual along the genome, and the bottom panel shows probability of seeing no ROH. (A) Simulation with no ascertainment, contamination, or ROH. (B) Simulation with ROH. (C) Simulation with ROH and ascertainment. (D) Simulation with ROH and contamination.}
    \label{fig2:ROH}
\end{figure}
We use a cutoff of 1 to compare our method performance to READ, while using a cutoff of 1 standard deviation for READ. Fig.\ref{fig4:Comparison_READ_KIN} shows this comparison using false positive and true positive rates for both the methods at different coverages under different simulations. We show that both methods have similar performance in the control case, and cases with ROH and ascertainment, except for $2^{nd}$ Degree relation, where KIN has higher power, especially at low coverages. We show that adding contamination adversely affects both the False Positve and True Positive rates corresponding to 1st and 2nd Degree relations for READ, while this has no effect on KIN performance.

@@

\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/kinplot.png}
    \centering
    \caption{Application of KIN on Neandertal specimens from Chagyrskaya. The color of a square represents the relatedness, while the number within denotes log likelihood ratio between the two maximum likelihood models.}
    \label{fig5:Chagyrskaya_KIN}
\end{figure}

We further applied KIN to a genome-wide dataset of 118 ancient individuals sequenced and analyzed by Mittnik et al.,2019. We compared our relatedness estimates to those obtained from the authors using READ and lcMLKIN, and found that in most cases the three methods agree. Out of 6903 pairs, 346 pairs show different relatedness for KIN and READ (table S2). However, there are only 31 pairs where KIN has $\Lambda >1$. We looked in detail at these pairs, and we found that it is possible to deduce the correct relatedness for several of these pairs by plotting the proportion of differences along the genome. In all the pairs that we analyzed in this way, KIN seems to predict the true relatedness. We show few examples of such cases in fig. S\ref{figS8:eg1}. We found 20 pairs in this subset of 346 pairs where lcMLKIN has sufficient data (Overlapping SNPs $>$ 10000). Among these 20 pairs, we found that lcMLKIN matches KIN estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLKIN predicts 3rd-5th degree, KIN predicts second degree) and pair AITI72-AITI77A (READ and lcMLKIN predict unrelated, while KIN estimates 3rd degree with log likelihood=1.7).  
We found 43 pairs for which READ estimates first degree, and compared the estimates of lcMLKIN and KIN for these cases (suppl table3). There were a total of 20 pairs where lcMLKIN had sufficient data, and 26 pairs where $loglikelihood ratio>2$ for KIN. Among the 19 overlapping pairs, we found four pairs where KIN and lcMLKIN differ, and a visual inspection shows that in these cases, KIN seems to have the correct estimate (fig. S\ref{figS9:eg2}).

@@

\begin{table}
\caption{\label{tab:Table 2}Model performance for ROH prediction}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
    \hline
    Coverage & Control(bias) & Control(var) & R(bias) & R(var) & RC(bias) & RC(var) & RA(bias) & RA(var)\\
    \hline
    4x & -0.007 & 0.006 & -0.081 & 0.036 & -0.083 & 0.036 & -0.091 & 0.037\\
    \hline
    0.2x & -0.003 & 0.003 & -0.068 & 0.03 & -0.068 & 0.026 & -0.076 & 0.029\\
    \hline
    0.1x & -0.003 & 0.002 & -0.049 & 0.021 & -0.05 & 0.021 & -0.057 & 0.022\\
    \hline
\end{tabular}
\label{table1}
\end{table}
